---
title: "Imputation"
author: Arend Lis
bibliography: references.bib
biblio-style: apalike
link-citation: yes
colorlinks: yes
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imputation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", message = TRUE, warning = FALSE
)
```


```{r setup}
library(PRONE)
```

# Load Data (TMT)

Here, we are directly working with the [SummarizedExperiment](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html) data. For more information on how to create the SummarizedExperiment from a proteomics data set, please refer to the ["Get Started"](PRONE.html) vignette.

The example TMT data set originates from [@biadglegne_mycobacterium_2022].

```{r load_real_tmt}
data("tuberculosis_TMT_se")
se <- tuberculosis_TMT_se
```

# Preprocessing 

As we have seen in the Preprocessing phase, that samples "1.HC_Pool1" and "1.HC_Pool2" have been removed from the data set due to their high amount of missing values (more than 80% of NAs per sample), before imputing the data we will here remove these two samples.

```{r}
se <- remove_samples_manually(se, "Label", c("1.HC_Pool1", "1.HC_Pool2"))
```


# Impute Data

Currently, there is only a mixed imputation method available in PRONE: k-nearest neighbor imputation for proteins with missing values at random and a left-shifted Gaussian distribution for proteins with missing values not at random. Imputation can be performed on a selection of normalized data sets using the "ain" parameter in the `impute_SE` function. The default is to impute all assays (ain = NULL). 

```{r impute}
se <- impute_se(se, ain = NULL)
```

# Session Info

```{r}
utils::sessionInfo()
```

# References
